Turbulence and Particle Acceleration in Giant Radio Halos: the Origin of Seed Electrons

### 2016-04-21 paper.tex
Added files via upload
merged paper.tex and added tau_D figure plus text and caption.

\begin{table}
  \caption{Thermal quantities and timescales for different spatial
    regions in a Coma like cluster.}
\begin{tabular}{l c  c c}
\hline
\hline
& & spatial regions & \\
 & $0.1\,\RH^{(2)}$ & $0.3\,\RH^{(2)}$ & $\RH^{(2)}$   \\
\hline
thermal quantities$^{(1)}$ & & & \\
$\rho$ [$10^{-27}$ g cm$^{-3}$] & 3.0 & 1.6 & 0.15 \\
$T$ [$10^{8}$ K] & 1.4 & 1.0 & 0.58 \\
\hline
timescales$^{(1)}$ & & & \\
$\tau_\rmn{D}$(\Mprimary)$^{(3)}$ [Gyr] & 0.45 & 0.44 & 0.39 \\
$\tau_\rmn{D}$(\Mstream)$^{(3)}$  [Gyr] & 0.50 & 0.47 & 0.34 \\
$\tau_\rmn{D}$(\Mflatturb)$^{(3)}$  [Gyr] & 0.69 & 0.56 & 0.27 \\
$\tau_\rmn{IC/sync}(P=10^4\,m_ec)^{(4)}$ [Gyr] & 0.11 & 0.15 & 0.22 \\
$\tau_\rmn{had}(P=100\,m_pc)^{(5)}$ [Gyr] & 2.4 & 4.5 & 47 \\
$\tau_\rmn{coul}(P=\,m_ec)^{(6)}$  [Gyr] & 0.0092 & 0.017 & 0.17 \\
\hline
\end{tabular}
\begin{quote}
 Notes: \\ 
 (1) Median quantities from our simulated post-merging cluster g72a.\\
 (2) Radius of the giant radio halo in Coma where $\RH\approx0.6\,R_{200}$.\\
 (3) Fermi-II reacceleration for both electrons and protons at all energies.\\
 (4) Inverse Compton and synchrotron cooling for electrons.\\
 (5) Catastrophic losses for protons.\\
 (6) Coulomb cooling for electrons (protons factor $m_e/m_p$ smaller).\\
 \label{tab:timescales}
  \end{quote}
\end{table}

\subsection{Scaling relations for turbulent reacceleration}
To understand the radial dependence of the diffusion constant we
investigate how $D_\rmn{pp}$ scales with different parameters. This is
crucial in order to identify important parameters, and understand the
profile of reaccelerated CRs and radio emission. Combining
Eqns.~\ref{eq:dpp}-\ref{eq:Gamma_e} we find that $D_\rmn{pp}\propto
I_0/\rho c_s \propto
X_\rmn{tu}^2k_0\epsilon_\rmn{th}^{\alpha_\rmn{tu}-1}\sqrt{T}$. Here we
immediately see that if the turbulent profile trace the thermal energy
(i.e. $\alpha_\rmn{tu}=1$), the square root of temperature alone
determines the diffusion constant. We also see that $D_\rmn{pp}$ is
much more sensitive to $X_\rmn{tu}$ than $k_0$.

During reacceleration the CRps approximately gain a factor 
\begin{equation}
  \label{eq:Creacc}
  \Delta C_\rmn{reacc.} \sim
  \exp{\left[\frac{(2+\alpha)(\alpha+1)}{4}\frac{\tau_\rmn{cl}}{\tau_\rmn{D}}\right]}\,,
\end{equation}
if injection and cooling are neglected. The reaccelerated CRs are then
exponentially dependent on $D_\rmn{pp}$ as long as $\tau_D \lesssim
\tau_\rmn{cl}$, where $\tau_\rmn{cl}$ represent the duration of
reacceleration of the merger which we assume to be 650~Myrs long.
Note that a fixed reacceleration timescale $\tau_D\sim
1/D_\rmn{pp}\sim 0.2-0.7\,\rmn{Gyr}$ is required to explain the
observations at each radius and for each model (see
Table~\ref{tab:timescales}). This allows us to interchange
$\tau_\rmn{cl} \sim X_\rmn{tu}^2 \sim k_0$, i.e. a longer
reacceleration duration have to be compensated by lowering the level
of turbulence or increasing the physical injection scale of turbulence
to keep $\tau_D$ constant.

@@

% --- section: Analytical Results and discussion --- %
\section{Parameter space exploration with spherical models}
\label{sect:param_comp}
In this paper we rely on several critical parameters describing
relatively unknown non-thermal physics in the ICM. Here we try to
explore the robustness of our models to the most critical parameters
related to turbulence and CRs. Figure~\ref{fig:param_comp} shows their
impact on radio emission. We find that the level of turbulence
($X_\rmn{tu}$) has a large impact on both radio spectra and profiles,
while the parameter $\alpha_\rmn{tu}$ that determines the spatial
profile of turbulence mainly influence the radio profiles. As
expected, the radio profiles flatten significantly with a flatter
turbulent profile. A higher amount of turbulent energy also flatten
the radio profile, where the reason being that the steep profiles of
injected CRs have less impact on the radio emission for an increasing
reacceleration efficiency. Interestingly, assumptions about CRs have
less impact on radio emission. This can be understood by looking at
the scaling of parameters to the reacceleration timescale, where
$\tau_\rmn{D} \propto 1/D_\rmn{pp} \propto 1/X_\rmn{tu}^2$. From
Eqn.~\ref{eq:Creacc} we see that the CRps with a spectral index of
$\alpha_\rmn{CR}\approx2.5$ approximately gain a factor $\Delta
C_\rmn{reacc.} \sim
\exp{\left[1.7\frac{\tau_\rmn{cl}}{\tau_\rmn{D}}\right]}$ during
reacceleration if injection and cooling are neglected. The CRes gain a
similar factor from reacceleration, however in the region around
$0.1\,R_{200}$ the Coulomb cooling timescale
($\tau_\rmn{coul}\approx0.01\,\rmn{Gyr}$) is much shorter than
$\tau_\rmn{D}\approx0.5\,\rmn{Gyr}$, hence $\Delta C_\rmn{reacc.}$
will be greatly reduced for the electrons. The spatial region is
important when we derive the radio spectrum, since in order to compare
to observations we integrate the CR profile out to $\sim
0.2\,R_{200}$. For $X_\rmn{tu}=0.2$ and $\alpha_\rmn{tu}=0.8$ we find
that $\Delta C_\rmn{reacc.} \sim 10$. This is confirmed by the top
panels in Fig.~\ref{fig:param_comp}. Increasing the level of
turbulence to $X_\rmn{tu}=0.3$, gives us a $\Delta C_\rmn{reacc.}
\sim 100$, which is roughly confirmed by same figures. The factor 2-3
difference in the figure is due to the shorter reacceleration
timescale that increase the contribution from reaccelerated CRes that
would otherwise cool away on a much shorter timescale. Hence we
conclude that the radio emission is much more sensitive to the
exponential dependence on compressible turbulence than the linear
dependence on injected CRs.

We can also derive a lower critical level of turbulence where
reacceleration is negligible, while it rises sharply above this
threshold. We define the threshold by $\Delta C_\rmn{reacc.}  \lesssim
2$, which gives us an $X_\rmn{thres,tu} = \left(\frac{0.04\,\log(2)
}{1.7}\frac{\tau_\rmn{D}}{\tau_\rmn{cl}}\right)^{0.5} \approx 0.08$
for $\tau_\rmn{D} = 0.5\,\rmn{Myr}$. This threshold is also seen in
the top panels of figure~\ref{fig:param_comp}. For instance, for
$X_\rmn{tu}\gtrsim0.08$, a factor two change in $X_\rmn{tu}$ changes
the radio surface brightness by a factor $\sim 10-100$. In principle
this threshold behavior allows us to provide a strict lower bound on
turbulence in radio halos. An even more stringent bound on
$X_\rmn{tu}$ can be inferred from combining radio observations with
the gamma-ray observations (see section~\ref{sect:radio_spec}).

>>>>

\begin{table}
  \caption{Thermal quantities and timescales for different spatial
    regions in a Coma like cluster.}
\begin{tabular}{l c  c c}
\hline
\hline
& & spatial regions & \\
 & $0.1\,\RH^{(2)}$ & $0.3\,\RH^{(2)}$ & $\RH^{(2)}$   \\
\hline
thermal quantities$^{(1)}$ & & & \\
$\rho$ [$10^{-27}$ g cm$^{-3}$] & 3.0 & 1.6 & 0.15 \\
$T$ [$10^{8}$ K] & 1.4 & 1.0 & 0.58 \\
\hline
timescales$^{(1)}$ & & & \\
$\tau_\rmn{D}$(\Mprimary)$^{(3)}$ [Gyr] & 0.45 & 0.44 & 0.39 \\
$\tau_\rmn{D}$(\Mstream)$^{(3)}$  [Gyr] & 0.50 & 0.47 & 0.34 \\
$\tau_\rmn{D}$(\Mflatturb)$^{(3)}$  [Gyr] & 0.69 & 0.56 & 0.27 \\
$\tau_\rmn{IC/sync}(P=10^4\,m_ec)^{(4)}$ [Gyr] & 0.11 & 0.15 & 0.22 \\
$\tau_\rmn{had}(P=100\,m_pc)^{(5)}$ [Gyr] & 2.4 & 4.5 & 47 \\
$\tau_\rmn{coul}(P=\,m_ec)^{(6)}$  [Gyr] & 0.0092 & 0.017 & 0.17 \\
\hline
\end{tabular}
\begin{quote}
 Notes: \\ 
 (1) Median quantities from our simulated post-merging cluster g72a derived 
 during last 300~Myrs in time. \\
 (2) Radius of the giant radio halo in Coma where $\RH\approx0.6\,R_{200}$.\\
 (3) Fermi-II reacceleration for both electrons and protons at all energies.\\
 (4) Inverse Compton and synchrotron cooling for electrons.\\
 (5) Catastrophic losses for protons.\\
 (6) Coulomb cooling for electrons (protons factor $m_e/m_p$ smaller).\\
 \label{tab:timescales}
  \end{quote}
\end{table}

\subsection{Scaling relations for turbulent reacceleration}
To understand the radial dependence of the diffusion constant we
investigate how $D_\rmn{pp}$ scales with different parameters. This is
crucial in order to identify important parameters, and understand the
profile of reaccelerated CRs and radio emission. Combining
Eqns.~\ref{eq:dpp}-\ref{eq:Gamma_e} we find that 
\begin{equation}
  \label{eq:Dpp_scaling}
  D_\rmn{pp}\propto I_0/\rho c_s \propto 
X_\rmn{tu}^2k_0\epsilon_\rmn{th}^{\alpha_\rmn{tu}-1}\sqrt{T}\,.
\end{equation}
Here we immediately see that if the turbulent profile trace the
thermal energy (i.e. $\alpha_\rmn{tu}=1$), the square root of
temperature alone determines the diffusion constant. We also see that
$D_\rmn{pp}$ is much more sensitive to $X_\rmn{tu}$ than $k_0$.

During reacceleration the CRps approximately gain a factor 
\begin{equation}
  \label{eq:Creacc}
  \Delta C_\rmn{reacc.} \sim
  \exp{\left[\frac{(2+\alpha)(\alpha+1)}{4}\frac{\tau_\rmn{cl}}{\tau_\rmn{D}}\right]}\,,
\end{equation}
if injection and cooling are neglected. The reaccelerated CRs are then
exponentially dependent on $tau_D\sim 1/D_\rmn{pp}$ as long as $\tau_D
\lesssim \tau_\rmn{cl}$, where $\tau_\rmn{cl}$ represent the duration
of reacceleration of the merger which we assume to be 650~Myrs long.

In Figure~\ref{fig:tauD} we show how the turbulent reacceleration
timescales in our three models scale with radius. As expected, the
\Mprimary model with $\alpha_\rmn{tu}=0.88$ has the flattest profile
with $\tau_D\approx0.4\,\rmn{Gyr}$, where the small dip at large
radius driven by the decrease in thermal energy density. The
\Mflatturb model has the the flattest turbulent profile parameterized
by the smaller $\alpha_\rmn{tu}$ which explains the steepest $\tau_D$
profile. Note that a fixed reacceleration timescale is required to
explain the observations at each radius and for each model (see also
Table~\ref{tab:timescales}). This allows us to interchange the value
of uncertain parameters $\tau_\rmn{cl} \sim X_\rmn{tu}^2 \sim k_0$,
i.e. a longer reacceleration duration have to be compensated by
lowering the level of turbulence or increasing the physical injection
scale of turbulence to keep $\tau_D$ constant.

\begin{figure}
  \includegraphics[width=1.0\columnwidth]{tau_reacc.eps}
  \caption{Turbulent reacceleration timescales for our simulated
    cluster g72a. We show in linear-log the reacceleration timescale
    ($\tau_D$) as a function of radius $R$ for our three models:
    \Mprimary (green dotted line), \Mstream (blue dashed line), and
    \Mflatturb (red solid line). Note that the timescales are derived
    during the last 100~Myrs in time in our simulations.}
  \label{fig:tauD}
\end{figure}

@@

% --- section: Analytical Results and discussion --- %
\section{Parameter space exploration with spherical models}
\label{sect:param_comp}
In this paper we rely on several critical parameters describing
relatively unknown non-thermal physics in the ICM. Here we try to
explore the robustness of our models to the most critical parameters
related to turbulence and CRs. Figure~\ref{fig:param_comp} shows their
impact on radio emission. We find that the level of turbulence
($X_\rmn{tu}$) has a large impact on both radio spectra and profiles,
while the parameter $\alpha_\rmn{tu}$ that determines the spatial
profile of turbulence mainly influence the radio profiles. As
expected, the radio profiles flatten significantly with a flatter
turbulent profile. A higher amount of turbulent energy also flatten
the radio profile, where the reason being that the steep profiles of
injected CRs have less impact on the radio emission for an increasing
reacceleration efficiency. Interestingly, assumptions about CRs have
less impact on radio emission. This can be understood by looking at
the scaling of parameters to the reacceleration timescale, where
$\tau_\rmn{D} \propto 1/D_\rmn{pp} \propto 1/X_\rmn{tu}^2$. From
Eqn.~\ref{eq:Creacc} we see that the CRps with a spectral index of
$\alpha_\rmn{CR}\approx2.5$ approximately gain a factor $\Delta
C_\rmn{reacc.} \sim
\exp{\left[1.7\frac{\tau_\rmn{cl}}{\tau_\rmn{D}}\right]}$ during
reacceleration if injection and cooling are neglected. The CRes gain a
similar factor from reacceleration, however in the region around
$0.1\,R_{200}$ the Coulomb cooling timescale
($\tau_\rmn{coul}\approx0.01\,\rmn{Gyr}$) is much shorter than
$\tau_\rmn{D}\approx0.6\,\rmn{Gyr}$, hence $\Delta C_\rmn{reacc.}$
will be greatly reduced for the electrons. The spatial region is
important when we derive the radio spectrum, since in order to compare
to observations we integrate the CR profile out to $\sim
0.2\,R_{200}$. For $X_\rmn{tu}=0.2$ and $\alpha_\rmn{tu}=0.8$ we find
that $\Delta C_\rmn{reacc.} \sim 10$. This is confirmed by the top
panels in Fig.~\ref{fig:param_comp}. Increasing the level of
turbulence to $X_\rmn{tu}=0.3$, gives us a $\Delta C_\rmn{reacc.}
\sim 100$, which is roughly confirmed by same figures. The factor 2-3
difference in the figure is due to the shorter reacceleration
timescale that increase the contribution from reaccelerated CRes that
would otherwise cool away on a much shorter timescale. Hence we
conclude that the radio emission is much more sensitive to the
exponential dependence on compressible turbulence than the linear
dependence on injected CRs.

We can also derive a lower critical level of turbulence where
reacceleration is negligible, while it rises sharply above this
threshold. We define the threshold by $\Delta C_\rmn{reacc.}  \lesssim
2$, which gives us an $X_\rmn{thres,tu} = \left(\frac{0.04\,\log(2)
}{1.7}\frac{\tau_\rmn{D}}{\tau_\rmn{cl}}\right)^{0.5} \approx 0.08$
for $\tau_\rmn{D} = 0.6\,\rmn{Gyr}$. This threshold is also seen in
the top panels of figure~\ref{fig:param_comp}. For instance, for
$X_\rmn{tu}\gtrsim0.08$, a factor two change in $X_\rmn{tu}$ changes
the radio surface brightness by a factor $\sim 10-100$. In principle
this threshold behavior allows us to provide a strict lower bound on
turbulence in radio halos. An even more stringent bound on
$X_\rmn{tu}$ can be inferred from combining radio observations with
the gamma-ray observations (see section~\ref{sect:radio_spec}).
